Partial and total correctness \\ as greatest and least fixed points

### 2015-01-21 paper.tex
+

\usepackage{MathUnicode}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{bm}

@@

\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}

@@

Section~\ref{sec:result} provides characterisations of partial and
total correctness, that differ only in that the former takes a
function's greatest fixed point where the latter takes its least, and
Section~\ref{sec:isabelle} describes how this characterisation is
proved correct using the Isabelle theorem
prover. Section~\ref{sec:infinite_branching} describes a condition
that is necessary for the fixed point characterisation of \emph{total}
correctness to be accurate; namely, that there is no infinite
branching. We begin, in Section~\ref{sec:preliminaries}, by
establishing some preliminary definitions.

@@

\paragraph{Modal-$μ$ calculus} We employ the following constructions
from the modal-$μ$ calculus~\cite{kozen83} to describe properties of our
transition relation. In the following, we suppose that
$p ∈ \pow(\Config)$ and that $\varphi:\pow(\Config) → \pow(\Config)$
is a monotone function.
\[\begin{array}{r@{~~}c@{~~}l}
\BOX p & = & \{C \mid ∀C'∈\Next(C)\ldotp C' ∈ p\} \\
\DIAMOND p & = & \{C \mid ∃C'∈\Next(C)\ldotp C' ∈ p\} \\
μX\ldotp \varphi(X) & = & \bigcap\{S\mid \varphi(S) ⊆ S\} \\
νX\ldotp \varphi(X) & = & \bigcup\{S\mid S ⊆ \varphi(S)\}
\end{array}\]

@@

\begin{definition} We characterise partial/total correctness (with
respect to postcondition $Q$) as the greatest/least fixed point of the
function $\SafeOne_Q$, defined as follows:
\[
\begin{array}{r@{~~}c@{~~}l}
\SafeOne_Q(X) &\eqdef& \{\config{c}{σ}\mid  \stuck{\config{c}{σ}} ⇒ σ ∈ Q\} ∩ \Box X.
\end{array}
\]
\end{definition}
%
We now establish a few properties of $\SafeOne_Q$. The first enables
the use of the greatest \emph{post}-fixed point and the least
\emph{pre}-fixed point of $\SafeOne_Q$ as proxies, respectively, for
its greatest and least fixed points. 
%
\begin{lemma}[Monotonicity]
\label{lem:mono}
%
$\SafeOne_Q$ is monotone; i.e., $X \subseteq X'$ implies
$\SafeOne_Q(X)\subseteq\SafeOne_Q(X')$.
\end{lemma}
%
The next lemmas allow $\SafeOne_Q$'s fixed points to be constructed
via series of approximants.
%
\begin{lemma}[GLB-preservation]
\label{lem:glb-pres}
%
$\SafeOne_Q$ preserves greatest lower bounds. That is, for any
ascending chain $x_0\subseteq x_1 \subseteq \dots$, we have
$\SafeOne_Q(\bigcap_k x_k) = \bigcap_k \SafeOne_Q(x_k)$.
%
\end{lemma}
%
\begin{lemma}[LUB-preservation]
\label{lem:lub-pres}
%
If our transition system has finite branching, then $\SafeOne_Q$
preserves least upper bounds. That is, if $\finite(\Next(C))$ holds
for all $C$, then for any ascending chain
$x_0\subseteq x_1 \subseteq \dots$, we have
$\SafeOne_Q(\bigcap_k x_k) = \bigcap_k \SafeOne_Q(x_k)$.
%
\end{lemma}
%
The following lemma states that in the absence of infinite branching,
every always-terminating command has a longest trace.
%
\begin{lemma}[Longest trace]
\label{lem:longest-trace}
%
If $\Next(C)$ is finite for all $C$, and $C_0 ∈ \terminates$, then
there exists an upper bound $M$ for which
$∀π ∈ \tracesfrom(C_0)．∃j≤M．\length(π) = j$.
%
\end{lemma}
%
\begin{proof} The lemma holds because $\tracesfrom(C_0)$ is a finite
set; this observation in turn rests on K\H{o}nig's infinity lemma.
\end{proof}

\begin{theorem}
\label{thm:main}
%
Partial and total correctness can be characterised as greatest/least
fixed points. Note that the third implication below relies
on our transition system having the property of finite branching.
%
\begin{eqnarray}
\label{thm1_pc} 
\seqspec{P}c{Q} & = & (\{c\}×P) ⊆ νX\ldotp\SafeOne_Q(X) \\
\label{thm1_tc1} 
\totalspec{P}c{Q} & ⇒ & (\{c\}×P) ⊆ μX\ldotp\SafeOne_Q(X) \\
\label{thm1_tc2}
\totalspec{P}c{Q} & ~⇐~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X)\quad
\text{if $\finite(\Next(C))$ for all $C$.}
\end{eqnarray}
%
\end{theorem}

\begin{proof}
We begin with the fixed point characterisation of partial
correctness. To prove \eqref{thm1_pc}, it suffices to prove that
$\Safe_Q$ coincides with $\SafeOne_Q$'s greatest post-fixed point.
\begin{eqnarray}
\label{eq:lem_pc}
\Safe_Q &=& νX\ldotp \SafeOne_Q(X).
\end{eqnarray}
%
For the $(⊆)$ direction of~\eqref{eq:lem_pc}, it is straightforward to
show, by expanding definitions and invoking standard lemmas about
reflexive transitive closures, that $\Safe_Q$ is a post-fixed point
(that is, $\Safe_Q ⊆ \SafeOne_Q (\Safe_Q)$), and hence that it is
below the \emph{greatest} post-fixed point. To show the $(⊇)$
direction, we first observe that we can invoke Lemma~\ref{lem:glb-pres}
to construct $\SafeOne_Q$'s greatest post-fixed point as the
intersection of a sequence of approximants:
%
\begin{eqnarray*}
νX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcap_k\left(\SafeOne_Q^k(\Config)\right).
\end{eqnarray*}
(The intuition is that the $k$th approximant only contains
configurations for which every trace that terminates in fewer than $k$
steps satisfies the postcondition when it does so.) It then suffices to
prove that for all $k$ we have
\[
∀C ∈ \SafeOne_Q^k (\Config). ∀π ∈ \tracesfrom(C)．∀j<k．\length(π) = j+1 ⇒ \snd (π_j) ∈ Q,
\] 
a fact that can be dispatched via mathematical induction on $k$.

We now turn to the fixed point characterisation of total correctness,
which is proved using roughly symmetric reasoning. To
prove~\eqref{thm1_tc1}, it suffices to show
that
\begin{eqnarray*}
\Safe_Q ∩ \terminates &⊇& μX\ldotp \SafeOne_Q(X),
\end{eqnarray*}
%
which follows from $\Safe_Q ∩ \terminates$ being a pre-fixed point of
$\SafeOne_Q$ and hence above its \emph{least} pre-fixed point. To
show~\eqref{thm1_tc2}, we first use Lemma~\ref{lem:lub-pres} to equate
the least pre-fixed point to the union of a sequence of approximants
as follows:
%
\begin{eqnarray*}
μX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcup_n\left(\SafeOne_Q^n(∅)\right).
\end{eqnarray*}
%
(The intuition is that the $k$th approximant contains all
configurations for which every trace terminates in fewer than $k$
steps and satisfies the postcondition when it does so.) We then show,
by mathematical induction on $k$ as before, that
%
\[
∀C．(∀π ∈ \tracesfrom(C)． ∃j < k．\length(π) = j+1 ∧ \snd (π_j) ∈ Q) ⇒ C ∈ \SafeOne_Q^{k+1}(\emptyset),
\]
which completes the proof. \qed
\end{proof}

\section{Proof mechanisation}
\label{sec:isabelle}

The proof presented in the previous section has been formalised in the
Isabelle theorem prover, with the exception of the following lemma,
which has currently only been proved `by hand'.

@@

If the $\trans$ relation allows infinite branching -- that is, if a
configuration can have infinitely many next configurations to choose
from -- then the least fixed point calculation does not coincide with
total correctness. The technical reason for the failure is that
$\SafeOne_Q$ no longer preserves least upper bounds. Intuitively, the
failure can be explained by the following counterexample.
\[
\begin{tikzpicture}
\node[draw] (c0) at (3,0) {$C_0$};
\node[draw] (c21) at (1,1) {$C_1$};
\node[draw] (c22) at (1,2) {\phantom{$C_0$}};
\node[draw] (c31) at (2,1) {$C_2$};
\node[draw] (c32) at (2,2) {\phantom{$C_0$}};
\node[draw] (c33) at (2,3) {\phantom{$C_0$}};
\node[draw] (c11) at (3,1) {$C_3$};
\node[draw] (c12) at (3,2) {\phantom{$C_0$}};
\node[draw] (c13) at (3,3) {\phantom{$C_0$}};
\node[draw] (c14) at (3,4) {\phantom{$C_0$}};
\node             at (4,1) {$\dots$};
\node[draw] (c41) at (5,1) {$C_k$};
\node[draw] (c42) at (5,2) {\phantom{$C_0$}};
\node       (c43) at (5,3) {$\vdots$};
\node[draw] (c44) at (5,4) {\phantom{$C_0$}};
\node             at (6,1) {$\dots$};
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c21);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c21) -- (c22);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c31);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c31) -- (c32);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c32) -- (c33);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c11);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c11) -- (c12);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c12) -- (c13);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c13) -- (c14);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c0) -- (c41);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c41) -- (c42);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c42) -- (c43);
\draw[line join=round,
decorate, ->,decoration={
    zigzag,
    segment length=4,
    amplitude=.9,post=lineto,
    post length=2pt
}] (c43) -- (c44);
\draw[decorate,decoration={brace,amplitude=10pt}]
([xshift=5mm]c44.north) -- ([xshift=5mm]c42.south) node [black,midway,xshift=6mm] {$k$};
\end{tikzpicture}
\]
The $k$th approximant contains configurations whose traces all
terminate in fewer than $k$ steps. No approximant can contain
configuration $C_0$, since there is no bound $k$ on the length of its
traces. Yet this configuration will be admitted by a total correctness
specification whose postcondition is $\mathit{true}$, since it is the
case that each of its traces terminates.

Therefore, in the presence of infinite branching, the fixed point
characterisations of total correctness must be weakened to an
implication. Preservation of greatest lower bounds is unaffected by
infinite branching, so the characterisation of partial correctness
remains intact.

\begin{eqnarray*}
\seqspec{P}c{Q} & ~⇔~ & (\{c\}×P) ⊆ νX\ldotp \SafeOne_Q(X) \\
\totalspec{P}c{Q} & ~⇐~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X) 
\end{eqnarray*}

@@

Clarke~\cite{clarke79} was the first to study the correspondence
between greatest/least fixed points and partial/total correctness. He
remarks that
\begin{quote}the fixedpoints of $Γ$ form a complete lattice under the
natural partial ordering on $\pow(\State)$. The top element of this
lattice is the weakest precondition for partial correctness and the
bottom element is the weakest precondition for total
correctness.~\cite[p.~279]{clarke79}
\end{quote} 
%
We note three ways in which the current work extends on Clarke's
original observation. First, Clarke's work is in the setting of a
particular programming language that provides syntax for sequencing,
conditionals, assignment and recursive calls to parameterless
procedures. In this paper, we need not tackle issues of programming
language syntax, since we work at the level of arbitrary transition
systems. Second, Clarke's work is in the setting of a big-step
semantics; that is, the meaning of each programming construct maps an
initial state directly to a final state. We work with a small-step
semantics (as described by a transition system), which means that our
approach handles parallel programs without any further work. Big-step
semantics is well-known to be unable to handle parallelism, as
acknowledged by Clarke, who remarks that he `[is] currently attempting
to extend this fixedpoint theory to additional programming features
such as parallelism'~\cite[p.~292]{clarke79}. The definition of total
correctness in a small-step setting requires more nuance than in a
big-step setting, where it is trivial. Third, Clarke's work applies
only in a deterministic setting: a program has either a single final
state, or none at all. We allow for highly non-deterministic programs,
and only restrict the amount of non-determinism in the case of total
correctness.

Recent work on various Hoare logics for concurrency have used (mildly
disguised) greatest fixed point calculations to obtain a partial
correctness semantics. See, for example,
\cite[Definition~3.2]{vafeiadis11} and
\cite[Definition~25]{dinsdale-young+13}. The result in this paper
could prove helpful when those program logics are extended to total
correctness.

Jacobs and Gries have proposed \emph{general correctness} as a way to
unify partial and total correctness~\cite{jacobs+85}. It would be
interesting to investigate whether general correctness can be also
characterised as a fixed point calculation.

The idea of using the least and greatest fixed point of the same
function has been previously exploited by
Paulson~\cite[\S3]{paulson97a}, who obtains the set of finite lists
from a function's least fixed point, and the set of possibly-infinite
(`lazy') lists from its greatest fixed point.

\paragraph{Acknowledgements} This work was supported by EPSRC grant
EP/K011499/1. I would like to thank Edmund Clarke, Alastair Donaldson,
Tony Hoare, Peter Lammich, and Andreas Lochbihler for helpful feedback
and discussions.

\bibliographystyle{abbrv}
\bibliography{/Users/jpw48/Dropbox/John}

>>>>

\usepackage{MathUnicode}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{framed}

@@

\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage{url}

@@

Section~\ref{sec:result} provides characterisations of partial and
total correctness, that differ only in that the former takes a
function's greatest fixed point where the latter takes its least.
Section~\ref{sec:infinite_branching} describes a condition that is
necessary for the fixed point characterisation of \emph{total}
correctness to be accurate; namely, that there is no infinite
branching. Section~\ref{sec:related} discusses related work, mainly
focussing on how our result extends a similar observation made by
Clarke in 1979~\cite{clarke79}. We begin, in
Section~\ref{sec:preliminaries}, by establishing some preliminary
definitions.

@@

\paragraph{Modal-$μ$ calculus} We employ the following constructions
from the modal-$μ$ calculus~\cite{kozen83} to describe properties of our
transition relation. In the following, we suppose that
$p ∈ \pow(\Config)$ and that $\varphi:\pow(\Config) → \pow(\Config)$
is a monotone function.
\[\begin{array}{r@{~~}c@{~~}l}
\BOX p & = & \{C．∀C'∈\Next(C)\ldotp C' ∈ p\} \\
\DIAMOND p & = & \{C． ∃C'∈\Next(C)\ldotp C' ∈ p\} \\
μX\ldotp \varphi(X) & = & \bigcap\{S．\varphi(S) ⊆ S\} \\
νX\ldotp \varphi(X) & = & \bigcup\{S．S ⊆ \varphi(S)\}
\end{array}\]

@@

\begin{definition}
\label{defn:safeone}
We characterise partial/total correctness (with
respect to postcondition $Q$) as the greatest/least fixed point of the
function $\SafeOne_Q$, defined as follows:
\[
\begin{array}{r@{~~}c@{~~}l}
\SafeOne_Q(X) &\eqdef& \{\config{c}{σ}．\stuck{\config{c}{σ}} ⇒ σ ∈ Q\} ∩ \Box X.
\end{array}
\]
\end{definition}
%
We now establish a few properties of $\SafeOne_Q$. The first enables
the use of the greatest \emph{post}-fixed point and the least
\emph{pre}-fixed point of $\SafeOne_Q$ as proxies, respectively, for
its greatest and least fixed points. 
%
\begin{lemma}[Monotonicity]
\label{lem:mono}
%
$\SafeOne_Q$ is monotone; i.e., $X \subseteq X'$ implies
$\SafeOne_Q(X)\subseteq\SafeOne_Q(X')$.
\end{lemma}
%
The next lemmas allow $\SafeOne_Q$'s fixed points to be constructed
via series of approximants.
%
\begin{lemma}[GLB-preservation]
\label{lem:glb-pres}
%
$\SafeOne_Q$ preserves greatest lower bounds. That is, for any
ascending chain $x_0\subseteq x_1 \subseteq \dots$, we have
$\SafeOne_Q(\bigcap_k x_k) = \bigcap_k \SafeOne_Q(x_k)$.
%
\end{lemma}
%
\begin{lemma}[LUB-preservation]
\label{lem:lub-pres}
%
If our transition system has finite branching, then $\SafeOne_Q$
preserves least upper bounds. That is, if $\finite(\Next(C))$ holds
for all $C$, then for any ascending chain
$x_0\subseteq x_1 \subseteq \dots$, we have
$\SafeOne_Q(\bigcap_k x_k) = \bigcap_k \SafeOne_Q(x_k)$.
%
\end{lemma}
%
The following lemma states that in the absence of infinite branching,
every always-terminating command has a longest trace.
%
\begin{lemma}[Longest trace]
\label{lem:longest-trace}
%
If $\Next(C)$ is finite for all $C$, and $C_0 ∈ \terminates$, then
there exists an upper bound $M$ for which
$∀π ∈ \tracesfrom(C_0)．∃j≤M．\length(π) = j$.
%
\end{lemma}
%
\begin{proof} We first recall K\H{o}nig's infinity lemma as it applies
to trees -- that every tree with infinitely-many vertices, each having
finitely-many successor vertices, has at least one infinite
trace~\cite{franchella97}:
\begin{equation}
\label{eq:konig}
¬\finite\{C'．C_0 \trans^* C'\} ⇒ (∀C．\finite(\Next(C))) ⇒
∃π∈\tracesfrom(C_0)．\dom(π) = \nat.
\end{equation}
Now, if $C_0∈\terminates$, then it has no infinite traces, and
hence, by the contrapositive of~\eqref{eq:konig}, the number
of configurations reachable from $C_0$ is finite. This number provides
a suitable upper bound $M$ on the length of traces from $C_0$. 
\end{proof}
%
\begin{framed}
\begin{theorem}
\label{thm:main}
%
Partial and total correctness can be characterised as greatest/least
fixed points. Note that the third implication below relies
on our transition system having the property of finite branching.
%
\begin{eqnarray}
\label{thm1_pc} 
\seqspec{P}c{Q} & = & (\{c\}×P) ⊆ νX\ldotp\SafeOne_Q(X) \\
\label{thm1_tc1} 
\totalspec{P}c{Q} & ⇐ & (\{c\}×P) ⊆ μX\ldotp\SafeOne_Q(X) \\
\label{thm1_tc2}
\totalspec{P}c{Q} & ~⇒~ & (\{c\}×P) ⊆ μX\ldotp \SafeOne_Q(X)\quad
\text{if $\finite(\Next(C))$ for all $C$.}
\end{eqnarray}
%
\end{theorem}
\end{framed}
%
\begin{proof}
We begin with the fixed point characterisation of partial
correctness. To prove \eqref{thm1_pc}, it suffices to prove that
$\Safe_Q$ coincides with $\SafeOne_Q$'s greatest post-fixed point.
\begin{eqnarray}
\label{eq:lem_pc}
\Safe_Q &=& νX\ldotp \SafeOne_Q(X).
\end{eqnarray}
%
For the $(⊆)$ direction of~\eqref{eq:lem_pc}, it is straightforward to
show, by expanding definitions and invoking standard lemmas about
reflexive transitive closures, that $\Safe_Q$ is a post-fixed point
(that is, $\Safe_Q ⊆ \SafeOne_Q (\Safe_Q)$), and hence that it is
below the \emph{greatest} post-fixed point. To show the $(⊇)$
direction, we first observe that we can invoke
Lemma~\ref{lem:glb-pres} to construct $\SafeOne_Q$'s greatest
post-fixed point as the intersection of a sequence of approximants:
%
\begin{eqnarray*}
νX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcap_k\left(\SafeOne_Q^k(\Config)\right).
\end{eqnarray*}
(The intuition is that approximant $k+1$ contains configurations for
which every stuck configuration that is reachable in $k$ steps
satisfies the postcondition.) After expanding the definition of
$\Safe$, it remains to show:
%
\begin{equation}
\label{eq:pc_mid}
(∀k．C ∈ \SafeOne_Q^k(\Config)) ⇒ ∀C'．((C \trans^* C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q.
\end{equation}
%
After rewriting $C \trans^* C'$ as $∃n．C \trans^n C'$ and then
instantiating the $k$ in~\eqref{eq:pc_mid} to $n+1$, it suffices to show, for all $n$:
\[
C ∈ \SafeOne_Q^{n+1}(\Config) ⇒ ∀C'．((C \trans^n C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q,
\]
which can be dispatched via mathematical induction on $n$ (with
$C$ universally quantified in the induction hypothesis).

We now turn to the fixed point characterisation of total correctness. To
prove~\eqref{thm1_tc1}, it suffices to show
that
\begin{eqnarray*}
\Safe_Q ∩ \terminates &⊇& μX\ldotp \SafeOne_Q(X),
\end{eqnarray*}
%
which follows from $\Safe_Q ∩ \terminates$ being a pre-fixed point of
$\SafeOne_Q$ and hence above its \emph{least} pre-fixed point. To
show~\eqref{thm1_tc2}, we first use Lemma~\ref{lem:lub-pres} to equate
the least pre-fixed point to the union of a sequence of approximants
as follows:
%
\begin{eqnarray*}
μX\ldotp \SafeOne_Q(X) &=& \textstyle\bigcup_k\left(\SafeOne_Q^k(∅)\right).
\end{eqnarray*}
%
(The intuition is that approximant $k+1$ contains all configurations
for which every trace terminates in no more than $k$ steps and satisfies
the postcondition when it does so.) After expanding the definition of
$\Safe$, and invoking Lemma~\ref{lem:longest-trace} to obtain the
length $M$ of $C$'s longest trace, it remains to show:
%
\begin{equation}
\label{eq:tc_mid}
\begin{array}{l}
(∃M．∀π ∈ \tracesfrom(C)．∃j≤M．\length(π) = j) ⇒ {}\\
(∀C'．((C \trans^* C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q) ⇒ (∃k．C ∈ \SafeOne_Q^k(\emptyset)).
\end{array}
\end{equation}
%
We choose $M$ as a witness for $k$ in~\eqref{eq:tc_mid}. It then
suffices to show, for all $M$:
%
\[
\begin{array}{l}
(∀π ∈ \tracesfrom(C)．∃j≤M．\length(π) = j) ⇒ {}\\
(∀C'．((C \trans^* C') ∧ \stuck{(C')}) ⇒ \snd(C') ∈ Q) ⇒ C ∈ \SafeOne_Q^M(\emptyset).
\end{array}
\]
%
With a little further algebraic manipulation we arrive at:
\[
(∀π ∈ \tracesfrom(C)．∃j≤M．\length(π) = j ∧ \snd(\pi_{j-1}) ∈ Q) ⇒ C ∈ \SafeOne_Q^M(\emptyset)
\]
which can be dispatched by mathematical induction on $M$, generalising
$C$ as before, thus completing the proof.\qed
\end{proof}

\paragraph{Proof mechanisation} Our main theorem and its accompanying
lemmas have been formalised and proved in the Isabelle theorem prover,
with the exception of Lemma~\ref{lem:longest-trace}, of which we
currently only have a hand proof. The proof relies on K\H{o}nig's
infinity lemma, which we found to be difficult to formalise. The
proof is available online.\footnote{\url{http://www.doc.ic.ac.uk/~jpw48/Partial_Total_Correctness_As_Fixed_Points.thy}}

@@

If the $\trans$ relation allows infinite branching -- that is, if a
configuration can have infinitely many next configurations to choose
from -- then the least fixed point calculation does not coincide with
total correctness. The technical reason for the failure is that
$\SafeOne_Q$ no longer preserves least upper bounds
(cf.~Lemma~\ref{lem:lub-pres}). Intuitively, the
failure can be explained by the following counterexample.
%
\[
\begin{tikzpicture}[zaggy/.style={line join=round,
decorate, ->,decoration={zigzag, segment length=4,
amplitude=.9,post=lineto, post length=2pt}}]
\node[draw] (c0) at (3,0) {$C_0$};
\node[draw] (c21) at (1,1) {$C_1$};
\node[draw] (c22) at (1,2) {\phantom{$C_0$}};
\node[draw] (c31) at (2,1) {$C_2$};
\node[draw] (c32) at (2,2) {\phantom{$C_0$}};
\node[draw] (c33) at (2,3) {\phantom{$C_0$}};
\node[draw] (c11) at (3,1) {$C_3$};
\node[draw] (c12) at (3,2) {\phantom{$C_0$}};
\node[draw] (c13) at (3,3) {\phantom{$C_0$}};
\node[draw] (c14) at (3,4) {\phantom{$C_0$}};
\node             at (4,1) {$\dots$};
\node[draw] (c41) at (5,1) {$C_k$};
\node[draw] (c42) at (5,2) {\phantom{$C_0$}};
\node       (c43) at (5,3) {$\vdots$};
\node[draw] (c44) at (5,4) {\phantom{$C_0$}};
\node             at (6,1) {$\dots$};
\draw[zaggy] (c0) to [in=270,out=180] (c21);
\draw[zaggy] (c21) -- (c22);
\draw[zaggy] (c0) to [in=270,out=180] (c31);
\draw[zaggy] (c31) -- (c32);
\draw[zaggy] (c32) -- (c33);
\draw[zaggy] (c0) -- (c11);
\draw[zaggy] (c11) -- (c12);
\draw[zaggy] (c12) -- (c13);
\draw[zaggy] (c13) -- (c14);
\draw[zaggy] (c0) to[in=270,out=0] (c41);
\draw[zaggy] (c41) -- (c42);
\draw[zaggy] (c42) -- (c43);
\draw[zaggy] (c43) -- (c44);
\draw[decorate,decoration={brace,amplitude=10pt}]
([xshift=5mm]c44.north) -- ([xshift=5mm]c42.south) node
[black,midway,xshift=6mm] {$k$};
\end{tikzpicture}
\]
Recall that approximant $k$ contains configurations whose traces all
terminate in fewer than $k$ steps. No approximant can contain
configuration $C_0$, since there is no bound on the length of its
traces. Yet this configuration will be admitted by a total correctness
specification whose postcondition is $\mathit{true}$, since it is the
case that each of its traces terminates.

Therefore, in the presence of infinite branching, the fixed point
characterisation of total correctness must be weakened to an
implication. Preservation of greatest lower bounds is unaffected by
infinite branching, so the characterisation of partial correctness
remains intact.

@@

Clarke~\cite{clarke79} was the first to study the correspondence
between greatest/least fixed points and partial/total correctness. He
remarks, in a 1979 article, that 
%
\begin{quote}the fixedpoints of $\Gamma$ form a complete
lattice under the natural partial ordering on $\pow(\State)$. The top
element of this lattice is the weakest precondition for partial
correctness and the bottom element is the weakest precondition for
total correctness.~\cite[p.~279, footnote]{clarke79} \end{quote}
%
We note three ways in which the current work extends on Clarke's
original observation. First, Clarke works in the setting of a
particular programming language that provides syntax for sequencing,
conditionals, assignment, and recursive calls to parameterless
procedures. In this paper, we need not tackle issues of programming
language syntax, since we work at the level of arbitrary transition
systems. Second, Clarke's work applies only in a deterministic
setting: a program has either a single final state, or none at all. We
allow for highly non-deterministic programs, and only restrict to
finite non-determinism in the case of total correctness. Third,
Clarke's work is in the setting of a big-step semantics; that is, the
meaning of each programming construct maps an initial state directly
to a final state. We work with a small-step semantics (as described by
a transition system), which means that our approach handles parallel
programs without further adaptation. Big-step semantics is well-known
to be unable to handle parallelism, as acknowledged by Clarke, who
remarks that he `[is] currently attempting to extend this fixedpoint
theory to additional programming features such as
parallelism'~\cite[p.~292]{clarke79}. The definition of total
correctness in a big-step and deterministic setting is trivial, and
hence Clarke's $\Gamma$ functional is straightforward -- this perhaps
explains why Clarke consigned the observation we quote above to a mere
footnote. Our functional, on the other hand, which we call $\SafeOne$,
is fairly subtle: experience shows that even minor modifications of
Definition~\ref{defn:safeone} quickly lead to either the least or the
greatest fixed point becoming degenerate.

Regarding other influences on this work: the idea of using the least
and greatest fixed point of the same function has been previously
exploited by Paulson~\cite[\S3]{paulson97a}, who obtains the set of
finite lists from a function's least fixed point, and the set of
possibly-infinite (`lazy') lists from its greatest fixed point.

Recent work on various Hoare logics for concurrency provide avenues
for further development of the current work. These logics typically
use (mildly disguised) greatest fixed point calculations to obtain a
partial correctness semantics; see, for example,
\cite[Definition~3.2]{vafeiadis11} and
\cite[Definition~25]{dinsdale-young+13}. If these logics were extended
to handle total correctness, the result presented in this paper could
ease the transition.

Another direction for future work is provided by Jacobs and Gries, who
have proposed \emph{general correctness} as a way to unify partial and
total correctness~\cite{jacobs+85}. It would be interesting to
investigate whether general correctness can be also characterised as a
fixed point calculation.
\[
\text{---}
\]
\paragraph{Acknowledgements} This work was supported by EPSRC grant
EP/K011499/1. I would like to thank Edmund Clarke, Alastair Donaldson,
Tony Hoare, Peter Lammich, and Andreas Lochbihler for helpful feedback
and discussions.

\newpage
\bibliographystyle{abbrv}
\bibliography{/Users/jpw48/Dropbox/John}
